Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-23 submission/Main_blinded.tex
submission materials

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  pdfauthor={Equipo EDUMER},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Perceptions of Inequality and Meritocracy: Their Interplay in
Shaping Preferences for Market Justice in Chile (2016-2023)}

  \author{Juan Carlos Castillo}
            \affil{%
                  Departamento de Sociología, Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Andreas Laffert}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
        \author{Julio Iturra-Sanhueza}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study investigates the relationship between perceptions of economic
inequality, meritocratic beliefs, and preferences for market justice in
Chile between 2016 and 2023. Using six waves of panel data from the
Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
8,643; \(N_{individuals}\) = 1,687), the analysis examines how
subjective assessments of inequality shape attitudes toward the role of
merit in access to key social services such as healthcare, education,
and pensions. Results indicate that rising perceptions of inequality are
associated with greater support for redistributive policies; however,
individuals with strong meritocratic convictions are more likely to
legitimize existing disparities. The study also considers the influence
of major social movements during this period, which appear to have
reshaped public discourse and perceptions of fairness. These findings
contribute to a deeper understanding of how beliefs about justice and
equity evolve in contexts marked by persistent inequality and entrenched
market-oriented frameworks. \newline \textbf{Keywords}: Economic
inequality, meritocracy, market justice, Chile, public preferences,
inequality perception
\end{abstract}

\setstretch{1.15}
\section{Introduction}\label{introduction}

Since 1980, economic inequality and wealth concentration have
dramatically increased worldwide, becoming one of the main challenges
for the social sciences. Globally, in 2021, less than 50\% of the world
population owned only 2\% of the wealth, while the richest 10\%
concentrated 76\%, and the wealthiest 1\% captured nearly 38\% of total
assets (\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This
context of economic disparity has sparked renewed interest in studying
not only the objective aspects of inequality, such as income and access
to resources, but also its subjective dimensions, including perceptions,
beliefs, and associated attitudes
(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). This has led,
among other reasons, to widespread concern in the social sciences about
the political delegitimization that economic inequality can produce
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
Understanding how people perceive inequality is crucial, as these
perceptions can influence how societies comprehend and justify the
distribution of goods and services.

The perception of economic inequality can be understood as an
individual's subjective assessment of how resources are allocated among
members of a given society (\citeproc{ref-akyelken_urban_2020}{Akyelken,
2020}). Regardless of their measurement, various studies have shown that
this perception often underestimates the gap between the rich and the
poor, which could have implications for attitudes toward the
distribution of goods and services
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-schroder_income_2017}{Schröder, 2017}). Moreover, in
recent years, there has been a discussion about the subjective and
objective aspects of inequality, demonstrating inconsistencies between
the two (\citeproc{ref-trump_income_2018}{Trump, 2018}). This tension is
relevant because advancing the study of perceptions could help
understand how perceived inequality affects the distribution of
resources within a society (Becker, 2020).

Perception of economic inequality has been associated with important
outcomes such as redistributive preferences, justification of
inequality, and legitimacy of the economic system
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). Research
shows that greater perceived inequality fosters stronger preferences for
redistribution, regardless of actual inequality levels
(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}). However, other findings suggest that heightened
perceptions of income gaps between high- and low-paying occupations may
lead to greater justification of inequality
(\citeproc{ref-Castillo2011}{Castillo, 2011}). A less explored area in
this regard concerns market justice, which denotes the degree to which
individuals consider just that social goods and services (e.g., health,
pensions, education) are allocated according to individual contribution,
competition, and ability to pay
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lane_market_1986}{Lane, 1986};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Market justice attitudes
reflect the belief that the market promotes procedural
fairness---equality of opportunity---so that outcomes result from
individual achievement (\citeproc{ref-lane_market_1986}{Lane, 1986}).
Indeed, Lindh (\citeproc{ref-lindh_public_2015}{2015}) shows that market
justice preferences are stronger in countries with higher private
spending on services than in those with more comprehensive welfare
systems. From this perspective, perceptions of economic inequality may
be key, as lower perceived gaps can reinforce acceptance of free-market
systems and justify existing inequalities.

Meritocracy posits that inequality can be legitimized through
distributive criteria such as effort and talent
(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
demonstrated that people with stronger meritocratic beliefs tend to
perceive less inequality as they attribute economic differences to
individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
inequality as these beliefs are associated with attitudes that
legitimize status differences
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
like Chile, where the distribution of goods and services is
predominantly governed by market logics strongly introduced during the
military dictatorship (1973-1989)
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}), these beliefs can
play a crucial role in the acceptance of social inequalities.
Incorporating the perception of meritocracy in this literature allows
for an understanding of how attitudes toward inequality are shaped not
only by the perception of economic gaps, but also by beliefs about how
these gaps are justified. Furthermore, these beliefs are often
consolidated from an early age, reinforced by institutions promoting
values such as effort and individual skills to climb socially
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).

The perception of meritocracy and the perception of economic inequality
would interact intricately in shaping preferences for market justice. On
the one hand, a stronger perception of merit as a criterion for
distribution could minimize the perception of economic gaps, thus
justifying an unequal system
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). On the other hand, the
perception of inequality could moderate the impact of meritocracy, as a
higher perception of economic gaps might challenge the idea that these
are solely based on merit.

The primary objective of this study is to analyze the interplay between
perceptions of inequality and meritocracy and their joint influence on
preferences for market justice in Chile from 2016 to 2023, using
longitudinal data from the ELSOC survey. This interaction is expected to
provide a more comprehensive explanation of attitudes toward market
justice in a country characterized by high inequality and strong
free-market influence (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-flores_top_2020}{Flores et al., 2020}). Additionally, this
analysis seeks to elucidate how political and social
contingencies---such as the 2019 and 2022 social movements---might have
moderated these relationships by prompting more critical reflection on
the commodification of social services. Recognizing the temporal
dimension in shaping market justice preferences is essential, given that
such preferences are not static but are influenced by historical and
contextual factors that challenge or reaffirm prevailing social norms.
In this regard, variations in perceived economic inequality and
meritocracy over time can affect how individuals endorse market-based
approaches. Consequently, public opinion may shift from supporting
market justice to embracing redistributive policies aimed at mitigating
inequalities.

\section{Theoretical views on market justice, inequality perception, and
meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}

\subsection{The justification of market
inequality}\label{the-justification-of-market-inequality}

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts'', whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
legitimacy of market justice stems from the assumption that inequalities
are not only inevitable but fair---so long as the rules are transparent
and opportunities are open. In this way, market justice provides a moral
justification for inequality by framing it as a necessary outcome of
individual responsibility.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{times}
\usepackage{setspace}
\usepackage{iftex}

\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Perceptions of Inequality and Meritocracy: Their Interplay in
Shaping Preferences for Market Justice in Chile (2016-2023)}
\author{ }

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study investigates the relationship between perceptions of economic
inequality, meritocratic beliefs, and preferences for market justice in
Chile between 2016 and 2023. Using six waves of panel data from the
Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
8,643; \(N_{individuals}\) = 1,687), the analysis examines how
subjective assessments of inequality shape attitudes toward the role of
merit in access to key social services such as healthcare, education,
and pensions. Results indicate that rising perceptions of inequality are
associated with greater support for redistributive policies; however,
individuals with strong meritocratic convictions are more likely to
legitimize existing disparities. The study also considers the influence
of major social movements during this period, which appear to have
reshaped public discourse and perceptions of fairness. These findings
contribute to a deeper understanding of how beliefs about justice and
equity evolve in contexts marked by persistent inequality and entrenched
market-oriented frameworks. \newline \textbf{Keywords}: Economic
inequality, meritocracy, market justice, Chile, public preferences,
inequality perception
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi

\setstretch{1.15}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Since 1980, economic inequality and wealth concentration have
dramatically increased worldwide, becoming one of the main challenges
for the social sciences. Globally, in 2021, less than 50\% of the world
population owned only 2\% of the wealth, while the richest 10\%
concentrated 76\%, and the wealthiest 1\% captured nearly 38\% of total
assets (\protect\hyperlink{ref-chancel_world_2022}{Chancel et al.,
2022}). This context of economic disparity has sparked renewed interest
in studying not only the objective aspects of inequality, such as income
and access to resources, but also its subjective dimensions, including
perceptions, beliefs, and associated attitudes
(\protect\hyperlink{ref-janmaat_subjective_2013}{Janmaat, 2013}). This
has led, among other reasons, to widespread concern in the social
sciences about the political delegitimization that economic inequality
can produce (\protect\hyperlink{ref-castillo_perception_2022}{Castillo
et al., 2022}). Understanding how people perceive inequality is crucial,
as these perceptions can influence how societies comprehend and justify
the distribution of goods and services.

The perception of economic inequality can be understood as an
individual's subjective assessment of how resources are allocated among
members of a given society
(\protect\hyperlink{ref-akyelken_urban_2020}{Akyelken, 2020}).
Regardless of their measurement, various studies have shown that this
perception often underestimates the gap between the rich and the poor,
which could have implications for attitudes toward the distribution of
goods and services
(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
2022}; \protect\hyperlink{ref-schroder_income_2017}{Schröder, 2017}).
Moreover, in recent years, there has been a discussion about the
subjective and objective aspects of inequality, demonstrating
inconsistencies between the two
(\protect\hyperlink{ref-trump_income_2018}{Trump, 2018}). This tension
is relevant because advancing the study of perceptions could help
understand how perceived inequality affects the distribution of
resources within a society (Becker, 2020).

Perception of economic inequality has been associated with important
outcomes such as redistributive preferences, justification of
inequality, and legitimacy of the economic system
(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
2022};
\protect\hyperlink{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et
al., 2020},
\protect\hyperlink{ref-garcia-sanchez_perceptions_2018}{2018}). Research
shows that greater perceived inequality fosters stronger preferences for
redistribution, regardless of actual inequality levels
(\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
al., 2019};
\protect\hyperlink{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}). However, other findings suggest that heightened
perceptions of income gaps between high- and low-paying occupations may
lead to greater justification of inequality
(\protect\hyperlink{ref-Castillo2011}{Castillo, 2011}). A less explored
area in this regard concerns market justice, which denotes the degree to
which individuals consider just that social goods and services (e.g.,
health, pensions, education) are allocated according to individual
contribution, competition, and ability to pay
(\protect\hyperlink{ref-kluegel_legitimation_1999}{Kluegel et al.,
1999}; \protect\hyperlink{ref-lane_market_1986}{Lane, 1986};
\protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015}). Market justice
attitudes reflect the belief that the market promotes procedural
fairness---equality of opportunity---so that outcomes result from
individual achievement (\protect\hyperlink{ref-lane_market_1986}{Lane,
1986}). Indeed, Lindh (\protect\hyperlink{ref-lindh_public_2015}{2015})
shows that market justice preferences are stronger in countries with
higher private spending on services than in those with more
comprehensive welfare systems. From this perspective, perceptions of
economic inequality may be key, as lower perceived gaps can reinforce
acceptance of free-market systems and justify existing inequalities.

Meritocracy posits that inequality can be legitimized through
distributive criteria such as effort and talent
(\protect\hyperlink{ref-davis_principles_2001}{Davis \& Moore, 2001};
\protect\hyperlink{ref-young_rise_1962}{Young, 1962}). Previous studies
have demonstrated that people with stronger meritocratic beliefs tend to
perceive less inequality as they attribute economic differences to
individual achievements (\protect\hyperlink{ref-mijs_paradox_2021}{Mijs,
2021}; \protect\hyperlink{ref-wilson_role_2003}{Wilson, 2003}) and also
to justify more inequality as these beliefs are associated with
attitudes that legitimize status differences
(\protect\hyperlink{ref-batruch_belief_2023}{Batruch et al., 2023}). In
a context like Chile, where the distribution of goods and services is
predominantly governed by market logics strongly introduced during the
military dictatorship (1973-1989)
(\protect\hyperlink{ref-boccardo_30_2020}{Boccardo, 2020}), these
beliefs can play a crucial role in the acceptance of social
inequalities. Incorporating the perception of meritocracy in this
literature allows for an understanding of how attitudes toward
inequality are shaped not only by the perception of economic gaps, but
also by beliefs about how these gaps are justified. Furthermore, these
beliefs are often consolidated from an early age, reinforced by
institutions promoting values such as effort and individual skills to
climb socially
(\protect\hyperlink{ref-castillo_socialization_2024}{Castillo et al.,
2024}; \protect\hyperlink{ref-reynolds_perceptions_2014}{Reynolds \&
Xian, 2014}).

The perception of meritocracy and the perception of economic inequality
would interact intricately in shaping preferences for market justice. On
the one hand, a stronger perception of merit as a criterion for
distribution could minimize the perception of economic gaps, thus
justifying an unequal system
(\protect\hyperlink{ref-castillo_percepcion_2019}{Castillo et al.,
2012}; \protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021}). On the
other hand, the perception of inequality could moderate the impact of
meritocracy, as a higher perception of economic gaps might challenge the
idea that these are solely based on merit.

The primary objective of this study is to analyze the interplay between
perceptions of inequality and meritocracy and their joint influence on
preferences for market justice in Chile from 2016 to 2023, using
longitudinal data from the ELSOC survey. This interaction is expected to
provide a more comprehensive explanation of attitudes toward market
justice in a country characterized by high inequality and strong
free-market influence
(\protect\hyperlink{ref-boccardo_30_2020}{Boccardo, 2020};
\protect\hyperlink{ref-flores_top_2020}{Flores et al., 2020}).
Additionally, this analysis seeks to elucidate how political and social
contingencies---such as the 2019 and 2022 social movements---might have
moderated these relationships by prompting more critical reflection on
the commodification of social services. Recognizing the temporal
dimension in shaping market justice preferences is essential, given that
such preferences are not static but are influenced by historical and
contextual factors that challenge or reaffirm prevailing social norms.
In this regard, variations in perceived economic inequality and
meritocracy over time can affect how individuals endorse market-based
approaches. Consequently, public opinion may shift from supporting
market justice to embracing redistributive policies aimed at mitigating
inequalities.

\hypertarget{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}{%
\section{Theoretical views on market justice, inequality perception, and
meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}}

\hypertarget{the-justification-of-market-inequality}{%
\subsection{The justification of market
inequality}\label{the-justification-of-market-inequality}}

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane
(\protect\hyperlink{ref-lane_market_1986}{1986}), who makes a contrast
between market justice and political justice. The author defines market
justice as a system of ``earned deserts'', whereby individuals are seen
as deserving of a determined distributive outcome due to their effort
and skills. In contrast, political justice emphasizes principles of
equality and need, which are often represented by the welfare state
action through social policies. An important remark is that the
principles of market justice rely on the assumption that markets are
neutral and self-regulating arenas, where individuals are treated fairly
because they face the same formal rules of engagement
(\protect\hyperlink{ref-lane_market_1986}{Lane, 1986}). Consequently,
the legitimacy of market justice stems from the assumption that
inequalities are not only inevitable but fair---so long as the rules are
transparent and opportunities are open. In this way, market justice
provides a moral justification for inequality by framing it as a
necessary outcome of individual responsibility.

Empirical studies have shown different strategies for the study of
market justice preferences. A common approach in the literature is to
gauge attitudes toward the legitimacy of inequality in specific domains,
especially when linked to income differences. This can be traced to the
seminal work of Kluegel and Smith
(\protect\hyperlink{ref-kluegel_beliefs_1981}{1981}) who assessed the
normative foundations that explain public support for economic
inequality. Over time, this approach has been extended beyond income to
include other market-mediated outcomes, such as education, healthcare,
and/or pensions. For example, Von Dem Knesebeck et al.
(\protect\hyperlink{ref-vondemknesebeck_are_2016}{2016}) and Immergut
and Schneider (\protect\hyperlink{ref-immergut_it_2020}{2020}) examine
whether citizens consider it fair that individuals with higher incomes
can access better healthcare, while Lee and Stacey
(\protect\hyperlink{ref-lee_fairness_2023}{2023}) apply a similar method
in the context of education in Australia. These studies usually rely on
a survey item asking respondents to evaluate the fairness of
income-based access to welfare services, allowing for comparing justice
perceptions across different contexts. Similarly, comparative studies
have also considered these indicators to study the support for
market-based distribution in welfare systems
(\protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015};
\protect\hyperlink{ref-svallfors_political_2007}{Svallfors, 2007}). More
recently, Castillo et al.
(\protect\hyperlink{ref-castillo_socialization_2024}{2024}) introduced a
single-item composite measure of market justice to assess student
attitudes toward income-based access to education, healthcare, and
pensions in Chile. These empirical strategies all aim to capture the
extent to which individuals accept inequality when framed as a
reflection of market outcomes.
